---
title: "distdiffR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{distdiffR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
`distdiffR` is an R package for bivariate two-sample tests of distributional equality.

This R package provides a collection of bivariate two-sample tests for distributional equality. The tests make use of statistics between empirical cumulative distribution functions averaged across a series of rotations and or toroidal shifts of the pooled samples. The variety of tests with their respective parameters can be called from the main function `distdiffr()`. It takes as input two bivariate samples (not necessarily the same size) in the form of two-column matrices, referred to as `data1` and `data2`.

Below is an example using Fisher's Iris data to show test results when the null hypotheses are true (and both distributions are equivalent). This is done by randomly assigning all three of the species or iris to two samples. Since this is a bivariate test of distributional equality, only the first two independent variables are used.

```{r setup, eval=FALSE}
library(distdiffR)

seedNum <- 123
set.seed(seedNum)

data(iris)
# Randomly assign all three species to two samples
iris <- iris[sample.int(nrow(iris)), ]
data1 <- as.matrix(iris[1:75, -(3:5)])
data2 <- as.matrix(iris[76:150, -(3:5)])
# Note: Data inputs must be matrices

# Rotational test
output <- distdiffr(data1,
                    data2,
                    testType = "rotational",
                    seedNum = seedNum)
output$pval
```

When `testType = "rotational"`, McKinney and Symanzik's rotational modified Syrjala test[^1] is being employed.

[^1]: McKinney, E., Symanzik, J., 2019. Modifications of the Syrjala Test for Testing Spatial Distribution Differences Between Two Populations, In: 2019 JSM Proceedings. American Statistical Association, Alexandria, VA. pp. 2518-2530.

Similar results are also shown for the toroidal and combined (rotational and toroidal) modified Syrjala tests:

```{r nullTrue, eval=FALSE}
# Toroidal shift test with proportions of points
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    propPnts = 0.1,
                    seedNum = seedNum)
output$pval == 0.566

# Toroidal shift test with thresholds below pooled sample size
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    seedNum = seedNum)
output$pval == 0.477

# Toroidal shift test with thresholds above pooled sample size
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    shiftThrshld = 100,
                    seedNum = seedNum)
output$pval == 0.502

# Toroidal shift test with a number of shifts
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    numShifts = 8,
                    seedNum = seedNum)
output$pval == 0.648

# Combined rotational and toroidal shift test
output <- distdiffr(data1,
                    data2,
                    testType = "combined",
                    seedNum = seedNum)
output$pval == 0.307
```

For any of the above tests, the data input order is arbitrary, e.g.,

```{r order, eval=FALSE}
output1 <- distdiffr(data1,
                     data2,
                     testType = "toroidal",
                     seedNum = seedNum)

output2 <- distdiffr(data2,
                     data1,
                     testType = "toroidal",
                     seedNum = seedNum)

output1$pval == output2$pval
```

Below demonstrates test results when the null hypothesis is false (i.e., there exists some difference between the two distributions). This is shown by separating the two samples by the species setosa and virginica, respectively.

```{r nullFalse, eval=FALSE}
data(iris)
data1 <- as.matrix(iris[iris[5] == "setosa", -(3:5)])
data2 <- as.matrix(iris[iris[5] == "virginica", -(3:5)])

# Rotational test
output <- distdiffr(data1,
                    data2,
                    testType = "rotational",
                    seedNum = seedNum)
output$pval == 0.001

# Toroidal shift test with proportions of points
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    propPnts = 0.1,
                    seedNum = seedNum)
output$pval == 0.001

# Toroidal shift test with thresholds below pooled sample size
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    seedNum = seedNum)
output$pval == 0.001

# Toroidal shift test with thresholds above pooled sample size
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    shiftThrshld = 100,
                    seedNum = seedNum)
output$pval == 0.001

# Toroidal shift test with a number of shifts
output <- distdiffr(data1,
                    data2,
                    testType = "toroidal",
                    numShifts = 8,
                    seedNum = seedNum)
output$pval == 0.001

# Combined rotational and toroidal shift test
output <- distdiffr(data1,
                    data2,
                    testType = "combined",
                    seedNum = seedNum)
output$pval == 0.001

```

Another version of the test also exists when combining bivariate data from multiple sources into each of the two-samples, respectively, which treats each contribution equally. This can be called via the `grouped_distdiffr()` function.

